Theories of deliberated choice

### 2024-04-30 simple.tex
observably closed

A claim about types of individuals is a set $C^\text{types} \wsub P^{\allhists}$. Given such a set, define the associated claim $C \wsub \allleadstos$ as $C = (C^\text{types})^I = \set{{\allleadstog} \in \allleadstos \suchthat \forall i \in I: {\ileadstog} \in C^\text{types}}$.
\begin{lemma}
  \label{th:typesfalsifiable}
  A claim about types of individuals $C^\text{types} \wsub P^{\allhists}$ 
  satisfies $\forall {\ileadstog} \in P^{\allhists} \setminus C^\text{types}: \exists H \in \finite(\incr(\allhists)) \suchthat \forall {\ileadstog[j]} \in C^\text{types}: \exists \hist \in H \suchthat {\ileadstog}(\hist) \neq {\ileadstog[j]}(\hist)$,
  iff its associated claim is falsifiable or trivial.
\end{lemma}
\begin{proof}
  Pick any ${\allleadstog} \notin C$.
  By definition of $C$,
  $\exists i \in I \suchthat {{\ileadstog}} \notin C^\text{types}$.
  As $C$ holds, ${\ileadstog} \notin \set{{\ileadsto[j]} \suchthat j \in I}$ (\cref{th:typesholds}).
  Pick any $H$ such that $\forall {\ileadstog[j]} \in C^\text{types}: \exists \hist \in H \suchthat {\ileadstog}(\hist) \neq {\ileadstog[j]}(\hist)$.
  Define $\theta = \set{(\hist, i, {\ileadstog}(\hist)) \suchthat \hist \in H} \in \Theta \cap \powerset{\allleadstog}$.
  Then, $\theta \notin C^\wsub$, because $\forall {\ileadstog[j]} \in C^\text{types}: \theta \nsubseteq {\ileadstog[j]}$.

@@

\begin{remark}
  The opposite of \cref{th:suffFals} is the negation of $\forall {\ileadstog} \in P^{\allhists} \setminus \set{{\ileadsto[k]} \suchthat k \in I}: \exists H \in \finite(\incr(\allhists)) \suchthat \forall j \in I: \exists \hist \in H \suchthat {\ileadstog}(\hist) \neq {\ileadsto[j]}(\hist)$.
  Thus, for some ${\allleadstog} \in \allleadstos$ and some $i \in I$ with ${\ileadstog} \in P^{\allhists} \setminus \set{{\ileadsto[k]} \suchthat k \in I}$, we have that $\forall H \in \finite(\incr(\allhists)): \exists j \in I \suchthat \forall \hist \in H: {\ileadstog}(\hist) = {\ileadsto[j]}(\hist)$.

  TODO
  In that case, if restricting our focus to claims about types of individuals, we need to include the real types, then close under suitable indistinguishability relation, so as to include for example $i$.

  Define $S$ as the relations in the claim. Define $R(S)$ as the adequate closure: those that we can’t reject.
  Does $S$ include $R(S)$? If so, it is falsifiable. Otherwise, there is in $R(S)$ a relation that we can’t reject.
  Let $Impl$ be all impliciations in $S$, of the form: if $\hist$ goes to $p$, then $\hist'$ goes to $p'$ (in the above example: if $a$ goes to $\phi$ then $aa$ goes to $\phi$). Any relation that breaks one of these implications is rejectable. For example, sending $a$ to $\phi$ and $ab$ to $¬\phi$ is rejectable.We need to include all relations that break no implication.
  TODO check all this. 
\end{remark}
\begin{definition}
  Say that an attitude $g \in P^{\allhists}$ is indistinguishable from a set of attitudes $S \wsub P^{\allhists}$ iff $\forall H \in \finite(\incr(\allhists)): \exists f \in S \suchthat \restr{g}{H} = \restr{f}{H}$.

  Define the extension of a claim about types $C^\text{types} \wsub P^{\allhists}$ to its indistiguishible attitudes as the the claim about types $R(C^\text{types}) = \set{g \in P^{\allhists} \suchthat \forall H \in \finite(\incr(\allhists)): \exists f \in C^\text{types} \suchthat \restr{g}{H} = \restr{f}{H}}$.
  Note that $C^\text{types} \wsub R(C^\text{types})$. 
  Also note that $S_1 \wsub S_2 \implies R(S_1) \wsub R(S_2)$.

  Define the closure under indistinguishability of a claim about types $C^\text{types}$ as $R^\infty(C^\text{types})$, equivalently, as the smallest set $C^\text{types}_\text{ind}$ such that $R(C^\text{types}_\text{ind}) \wsub C^\text{types}_\text{ind}$.
\end{definition}
The following result follows from \cref{th:typesfalsifiable}.
\begin{theorem}
  If a claim about types of individuals $C^\text{types} \wsub P^{\allhists}$ is not closed under indistinguishability, then it is not falsifiable and not trivial.
\end{theorem}
\begin{theorem}
  Given any situation, if the claim related to the claim about types of individuals $R^\infty({\allleadsto}(I))$ is trivial, then there exists no claim related to a claim about types of individuals that holds and is falsifiable.
\end{theorem}
\begin{proof}
  Consider a claim related to a claim about types of individuals $S$ that holds and that is falsifiable or trivial.
  For it to hold, we need ${\allleadsto}(I) \wsub S$.
  From \cref{th:typesfalsifiable}, for it to be falsifiable or trivial, we need that $\forall g \in P^{\allhists} \setminus S: \exists H \in \finite(\incr(\allhists)) \suchthat \forall f \in S: \restr{g}{H} \neq \restr{f}{H}$. In other words, all $g$ must be distinguishable from $S$. In other words, $S$ must be closed under indistinguishability. Thus, we need $R^\infty(S) \wsub S$.
  Also, because $S_1 \wsub S_2 \implies R(S_1) \wsub R(S_2)$, we need
  $R^\infty({\allleadsto}(I)) \wsub R^\infty(S) \wsub S$.
  It follows that $S$ is trivial.
\end{proof}

\begin{conjecture}
  Given any situation, the claim about types of individuals $R^\infty({\allleadsto}(I))$ holds, is anonymous, and is falsifiable or trivial.
\end{conjecture}

\begin{theorem}[To be deleted, probably]
  Consider the attitude where all individuals are either constant to $\phi$ except that they map $\emptyset$ to anything, or constant to $¬\phi$ except that they map $\emptyset$ to anything, say, $(\phi, \phi)$, $(¬\phi, \phi)$, $(\phi, ¬\phi)$, $(¬\phi, ¬\phi)$.
  Consider a claim about types of individuals containing all those ones. Assume that for some attitude $i$ not in the claim, no $H$ permit to distinguish $i$ from the claim. Then from the lemma above, the claim is not falsifiable. So, we need to include all attitudes that no $H$ permit to distinguish from the claim. Iterating, this is the closure under the indistinguishability relation.
  So, the claim about types must include the real ones and be closed under the indistinguishability relation.
\end{theorem}
\begin{theorem}
  Consider the attitude above. Is there a claim that holds and is anonymous and falsifiable?
  Check \cref{ex:falsAnonIfBig}.
\end{theorem}

>>>>

A claim about types of individuals is a set $C^\text{types} \wsub P^{\allhists}$. Given such a set, define the associated claim $C \wsub \allleadstos$ as $C = (C^\text{types})^I = \set{{\allleadstog} \in \allleadstos \suchthat \forall i \in I: {\ileadstog} \in C^\text{types}}$.
\begin{definition}
  Say that an attitude $g \in P^{\allhists}$ is observably included in a set of attitudes $S \wsub P^{\allhists}$ iff $\forall H \in \finite(\incr(\allhists)): \exists f \in S \suchthat \restr{g}{H} = \restr{f}{H}$.

  Define the extension of a claim about types $C^\text{types} \wsub P^{\allhists}$ to its observably included attitudes as the the claim about types $R(C^\text{types}) = \set{g \in P^{\allhists} \suchthat \forall H \in \finite(\incr(\allhists)): \exists f \in C^\text{types} \suchthat \restr{g}{H} = \restr{f}{H}}$.
  Note that $C^\text{types} \wsub R(C^\text{types})$. 
  Also note that $S_1 \wsub S_2 \implies R(S_1) \wsub R(S_2)$.

  Define the closure under observable inclusion of a claim about types $C^\text{types}$ as $R^\infty(C^\text{types})$, equivalently, as the smallest set $C^\text{types}_\text{ind}$ such that $R(C^\text{types}_\text{ind}) \wsub C^\text{types}_\text{ind}$.

  Given a set of attitudes $S$ and $\hist \in \allhists$, define the equivalence relation $=_\hist$ on $S$ as $f =_\hist g$ iff $f(\hist) = g(\hist)$. This also defines the quotient set $S / =_\hist$. (For the readers used to that terminology: if defining $\beta_\hist \in P^S$ as $\beta_\hist(f) = f(\hist)$, then $=_\hist$ is the equivalence kernel of $\beta_\hist$ and $S / =_\hist$ is the coimage of $\beta_\hist$.)
  Similarly, given $S$ and $H \in \finite(\incr(\allhists))$, define the equivalence relation $=_H$ on $S$ as $f =_H g$ iff $\restr{f}{H} = \restr{g}{H}$. This also defines the quotient set $S / =_H$.
\end{definition}
\begin{remark}
  A few more, maybe not useful, definitions.
  A function $l: \allhists \to \powerset{P^{\allhists}}$ is decreasing iff $\forall \hist \wsub \histp: l(\histp) \wsub l(\hist)$.
  Let $S \wsub P^{\allhists}$ be a set of attitudes.
  Say that a function $l: \allhists \to \powerset{P^{\allhists}}$ is an $S$-point-labeling function iff $\forall \hist \in \allhists: l(\hist) \in S / =_\hist$.
  Say that a function $l: \allhists \to \powerset{P^{\allhists}}$ is an $S$-sequence-labeling function iff $\forall \hist \in \allhists: l(\hist) \in S / =_{\intvl{\hist}}$.
\end{remark}
Let $\intvl{\hist} = (\hist_i)_{i \in \intvl{0, \card{\hist}}} \in \finite(\incr(\allhists))$ denote the increasing sequence of paths from $\emptyset$ to $\hist$.
\begin{conjecture}
  The $S$-labeling function associated to an attitude $f \in P^{\allhists}$ is the function $l^S_f: \allhists \to \powerset{P^{\allhists}}$ defined as $l^S_f(\hist) = \set{g \in S \suchthat g(\hist) = f(\hist)}$.

  An attitude is observably included in a set of attitudes $S$ iff $\forall \hist \in \allhists: \cap_{\histp \wsub \hist} l^S_f(\histp) \neq \emptyset$.
\end{conjecture}
\begin{proof}
  Consider $f$ with $\forall \hist \in \allhists: \cap_{\histp \wsub \hist} l^S_f(\histp) \neq \emptyset$.
  Consider any $H \in \finite(\incr(\allhists))$.
  We know that $\cap_{\histp \wsub \hist}{l^S_f(\histp)} \neq \emptyset$, thus in particular, defining $\hist = \max H$ and observing that $H \wsub \set{\histp \suchthat \histp \wsub \hist}$, we have that $\cap_{\histp \in H} l^S_f(\histp) \neq \emptyset$.
  Pick any $g \in \cap_{\histp \in H} l^S_f(\histp)$. Thus, $g \in S$.
  Also, $\forall \hist \in H: g \in l^S_f(\hist)$, thus,
  $\forall \hist \in H: g(\hist) = f(\hist)$.

  First observe that given any $S \wsub P^{\allhists}$, $f \in P^{\allhists}$ and $\hist \in \allhists$, as $\histp \wsub \hist$ iff $\histp \in \intvl{\hist}$, we have that $\cap_{\histp \wsub \hist} l^S_f(\histp) = \set{g \in S \suchthat \forall \histp \in \intvl{\hist}: g(\histp) = f(\histp)}$.

  It follows that if $\forall \hist \in \allhists: \cap_{\histp \wsub \hist} l^S_f(\histp) \neq \emptyset$, then $\forall H \in \finite(\incr(\allhists))$: $\emptyset \neq \set{g \in S \suchthat \forall \histp \in \intvl{\max H}: g(\histp) = f(\histp)} \wsub \set{g \in S \suchthat \forall \histp \in H: g(\histp) = f(\histp)}$.

  Conversely, if $\forall H \in \finite(\incr(\allhists)): \set{g \in S \suchthat \forall \histp \in H: g(\histp) = f(\histp)} \neq \emptyset$, then $\forall \hist \in \allhists$, defining $H = \intvl{\hist}$, we have that $\cap_{\histp \wsub \hist} l^S_f(\histp) \neq \emptyset$.
\end{proof}
\begin{remark}
  The opposite of \cref{th:suffFals} is the negation of $\forall {\ileadstog} \in P^{\allhists} \setminus \set{{\ileadsto[k]} \suchthat k \in I}: \exists H \in \finite(\incr(\allhists)) \suchthat \forall j \in I: \exists \hist \in H \suchthat {\ileadstog}(\hist) \neq {\ileadsto[j]}(\hist)$.
  Thus, for some ${\allleadstog} \in \allleadstos$ and some $i \in I$ with ${\ileadstog} \in P^{\allhists} \setminus \set{{\ileadsto[k]} \suchthat k \in I}$, we have that $\forall H \in \finite(\incr(\allhists)): \exists j \in I \suchthat \forall \hist \in H: {\ileadstog}(\hist) = {\ileadsto[j]}(\hist)$.

  TODO
  In that case, if restricting our focus to claims about types of individuals, we need to include the real types, then close under suitable indistinguishability relation (observable inclusion), so as to include for example $i$.

  Define $S$ as the relations in the claim. Define $R(S)$ as the adequate closure: those that we can’t reject.
  Does $S$ include $R(S)$? If so, it is falsifiable. Otherwise, there is in $R(S)$ a relation that we can’t reject.
  Let $Impl$ be all impliciations in $S$, of the form: if $\hist$ goes to $p$, then $\hist'$ goes to $p'$ (in the above example: if $a$ goes to $\phi$ then $aa$ goes to $\phi$). Any relation that breaks one of these implications is rejectable. For example, sending $a$ to $\phi$ and $ab$ to $¬\phi$ is rejectable.We need to include all relations that break no implication.
  TODO check all this. 
\end{remark}
\begin{lemma}
  \label{th:typesfalsifiable}
  A claim about types of individuals $C^\text{types} \wsub P^{\allhists}$ 
  satisfies $\forall g \in P^{\allhists} \setminus C^\text{types}: \exists H \in \finite(\incr(\allhists)) \suchthat \forall f \in C^\text{types}: \restr{g}{H} \neq \restr{f}{H}$,
  iff its associated claim is falsifiable or trivial.
\end{lemma}
\begin{proof}
  Pick any ${\allleadstog} \notin C$.
  By definition of $C$,
  $\exists i \in I \suchthat {{\ileadstog}} \notin C^\text{types}$.
  As $C$ holds, ${\ileadstog} \notin \set{{\ileadsto[j]} \suchthat j \in I}$ (\cref{th:typesholds}).
  Pick any $H$ such that $\forall {\ileadstog[j]} \in C^\text{types}: \exists \hist \in H \suchthat {\ileadstog}(\hist) \neq {\ileadstog[j]}(\hist)$.
  Define $\theta = \set{(\hist, i, {\ileadstog}(\hist)) \suchthat \hist \in H} \in \Theta \cap \powerset{\allleadstog}$.
  Then, $\theta \notin C^\wsub$, because $\forall {\ileadstog[j]} \in C^\text{types}: \theta \nsubseteq {\ileadstog[j]}$.

@@

The following result follows from \cref{th:typesfalsifiable}.
\begin{theorem}
  If a claim about types of individuals $C^\text{types} \wsub P^{\allhists}$ is not closed under observable inclusion, then it is not falsifiable and not trivial.
\end{theorem}
\begin{theorem}
  Given any situation, if the claim related to the claim about types of individuals $R^\infty({\allleadsto}(I))$ is trivial, then there exists no claim related to a claim about types of individuals that holds and is falsifiable.
\end{theorem}
\begin{proof}
  Consider a claim related to a claim about types of individuals $S$ that holds and that is falsifiable or trivial.
  For it to hold, we need ${\allleadsto}(I) \wsub S$.
  From \cref{th:typesfalsifiable}, for it to be falsifiable or trivial, we need that $\forall g \in P^{\allhists} \setminus S: \exists H \in \finite(\incr(\allhists)) \suchthat \forall f \in S: \restr{g}{H} \neq \restr{f}{H}$. In other words, all $g$ must be observably excludable from $S$. In other words, $S$ must be closed under observable inclusion. Thus, we need $R^\infty(S) \wsub S$.
  Also, because $S_1 \wsub S_2 \implies R(S_1) \wsub R(S_2)$, we need
  $R^\infty({\allleadsto}(I)) \wsub R^\infty(S) \wsub S$.
  It follows that $S$ is trivial.
\end{proof}
\begin{theorem}
  Given any situation, the claim about types of individuals $R^\infty({\allleadsto}(I))$ holds, is anonymous, and is falsifiable or trivial.
\end{theorem}

\begin{theorem}[To be deleted, probably]
  Consider the attitude where all individuals are either constant to $\phi$ except that they map $\emptyset$ to anything, or constant to $¬\phi$ except that they map $\emptyset$ to anything, say, $(\phi, \phi)$, $(¬\phi, \phi)$, $(\phi, ¬\phi)$, $(¬\phi, ¬\phi)$.
  Consider a claim about types of individuals containing all those ones. Assume that for some attitude $i$ not in the claim, no $H$ permit to observably exclude $i$ from the claim. Then from the lemma above, the claim is not falsifiable. So, we need to include all attitudes that no $H$ permit to observably exclude from the claim. Iterating, this is the closure under the observable inclusion relation.
  So, the claim about types must include the real ones and be closed under observable inclusion.
\end{theorem}
\begin{theorem}
  Consider the attitude above. Is there a claim that holds and is anonymous and falsifiable?
  Check \cref{ex:falsAnonIfBig}.
\end{theorem}
